Determinism, Complexity, and Predictability in Computer Performance

### 2014-04-22 results.tex
Liz is done rewriting

In this section, we offer an empirical validation of the two key
conjectures introduced in Section \ref{sec:intro}, namely:

@@

\item that the weighted permutation entropy (WPE) of a noisy
  real-valued time series is correlated with prediction
  accuracy---i.e., that the predictable structure in a time-series
  data set can be quantified by its WPE.

@@

\item that the way information is generated and processed by a system
  is correlated with the effectiveness of a given predictor on
  time-series data from that system

@@

Figure~\ref{fig:wpe_vs_mase_best} shows the {\sl best} of these 45
predictions for each system: that is, the lowest error over all 15
trials and all three methods for each of the eight programs.  The WPE
is plotted against the corresponding MASE value here in order to bring
out the correlation between these two quantities.
\begin{figure}[htbp]
  \centering
  \includegraphics[width=\columnwidth]{figs/prediction_vs_entropy}
  \caption{Weighted permutation entropy vs. mean absolute scaled error
    (MASE) of the best prediction of each system.
% 
% (across all trials and all prediction methods) 
% 
The dashed line is a least-squares linear fit of all the points except
{\tt dgesdd$_1$}, which we have excluded for reasons explained in the
text.}
  \label{fig:wpe_vs_mase_best}
\end{figure}
The best-case prediction error for each of these eight systems is
roughly proportional to the weighted permutation entropy, which is
consistent with our first conjecture.  The dashed line in the figure,
a linear least-squares fit to these best-case errors\footnote{with the
  exception of {\tt dgesdd$_1$}, for reasons described below} captures
this rough proportionality.  This is not a formal result.  The three
methods used here were chosen to span the space of standard prediction
strategies, but they do not cover that space exhaustively.  Our goal
here is an empirical assessment of predictability and complexity, not
formal results about a ``best'' predictor for a given time series.
There may be other methods that produce lower MASE values than those
in Figure~\ref{fig:wpe_vs_mase_best}, but the sparseness of the points
in the upper-left and lower-right quadrants of this plot strongly
suggests that the underlying predictability of a time series is
inversely correlated with its WPE.  The rest of this section describes
our results in more detail, including the measures taken to assure
meaningful comparisons across methods, trials, and programs, and
elaborates on the meaning of the dashed line in the figure.

Figure~\ref{fig:wpe_vs_mase_all} shows WPE vs. MASE plots for all 360
experiments.
\begin{figure}
  \centering
%    \includegraphics[width=\columnwidth]{figs/ARIMA_prediction_vs_entropy}
%    \includegraphics[width=\columnwidth]{figs/ARIMA_prediction_vs_entropy}
%    \includegraphics[width=\columnwidth]{figs/LMA_prediction_vs_entropy}
  \includegraphics[width=\columnwidth]{figs/predictions_vs_entropy3}
\caption{WPE vs. MASE for all trials, methods, and systems. \svdone,
  \svdthree, and \svdfive are omitted from the top plot for scale
  reasons.
% 
% ; their MASE scores are $2.676 \pm 4.328$, $31.39 \pm 0.282$, and
%   $20.87 \pm 0.192$ respectively. 
% 
Numerical values, including means and standard deviations of the
errors, can be found in Table~\ref{tab:error}.  The dashed line is the
same as in the previous figure.
}
    \label{fig:wpe_vs_mase_all}
\end{figure} 
There are 15 points in each cluster, one for each trial.  (The points
in Figure~\ref{fig:wpe_vs_mase_best} are the leftmost of the points
for the corresponding system in any of the three plots in
Figure~\ref{fig:wpe_vs_mase_all}.)  The WPE values do not vary very
much across trials.  For most traces, the variance in MASE scores is
low as well, resulting in small, tight clusters.  In some
cases---ARIMA predictions of {\tt col\_major}, for instance---the
MASE variance is larger, which spreads out the clusters
horizontally.  The distribution of the cluster {\sl positions} is
quite different in the three plots.  The mean MASE scores of
predictions generated with the nonlinear LMA method are generally
closer to the dashed line; the ARIMA method clusters are more
widely spread, the \naive ~clusters even more so.  This geometric
validation of the second conjecture in this paper is discussed in more
details in the following paragraphs.

@@

The dashed line in the Figures is a heuristic that captures the
argument at the end of the previous paragraph.  It can allow
practitioners to recognize when a prediction method is not well
matched to the task at hand: that is, when the time series has more
predictive structure than the method is able to use.  When the
predictions are well below that line (as in the top and middle images
in Figure~\ref{fig:wpe_vs_mase_all}), one should try a different
method.  Conversely, the position of the LMA cluster---well to the
left of the other two, near the dashed line---reflects this method's
ability to capture and exploit the structure that is present in this
signal.

@@

\svdone behaves very differently than the other seven systems in this
study.  Though its weighted permutation entropy is very high ($0.957
\pm 0.016$), two of the three prediction methods do quite well on this
signal, yielding MASE scores of $0.827 \pm 0.076$ (LMA) and $0.714 \pm
0.075$ (ARIMA).  This pushes the corresponding clusters of points in
the bottom two images in Figure~\ref{fig:wpe_vs_mase_all} well above
the trend followed by the other seven signals.  The MASE scores of the
predictions that were produced by the \naive ~method for this system,
however, are highly inconsistent.  The majority of the blue
diamond-shaped points on the top plot in
Figure~\ref{fig:wpe_vs_mase_all} are clustered near a MASE score of
0.6.  In five of the 15 \svdone trials, however, there were step
changes in the signal.  The \naive ~method has a very difficult time
with signals like this, particularly if there are multiple step
changes.  This raised the MASE scores of these trials, pushing the
corresponding points to the right\footnote{This includes the cluster
  of three points near MASE $\approx 2.5$, as well as two points that
  are beyond the domain of the graph, at MASE 11.2 and 14.8.}, and in
turn raising both the mean and variance of this set of trials.

The effects described in the previous paragraph are exacerbated by the
way MASE is calculated.  Recall that MASE scores are scaled
\emph{relative to a random-walk forecast}, which uses the current
value as the prediction.  This strategy works very badly on signals
with frequent, large, rapid transitions.  Consider a signal that
oscillates from one end of its range to the other at every step.  A
signal like this will have a low WPE, much like {\tt col\_major}.
However, a random-walk forecast of this signal will be 180 degrees out
of phase with the true continuation.  Since random-walk error appears
in the denominator of the MASE score, this effect can shift points
leftwards on a WPE vs. MASE plot, and that is exactly why the
\svdone clusters for ARIMA and LMA are above the dashed line in
Figure~\ref{fig:wpe_vs_mase_all}.  This time series, which is shown in
closeup in Figure~\ref{fig:svdone-ts},
\begin{figure}[htbp]
  \centering
    \includegraphics[width=\columnwidth]{figs/svdonets2}
\caption{A small portion of the \svdone time series}\label{fig:svdone-ts}
\end{figure} 
is not quite to the level of the worst-case signal described above,
but it still poses a serious challenge to random-walk prediction.  It
is dominated by a noisy regime (between $\approx$1.86 and
$\approx$1.88 on the vertical scale in the Figure), punctuated by
short excursions above 1.9.  In the former regime, which makes up more
than 80\% of the signal, there are frequent dips to 1.82 and
occasional larger dips below 1.8.  These single-point dips are the
bane of random-walk forecasting.  In this particular case, roughly
40\% of the forecasted points are off by the width of the associated
dip, which skews the associated MASE scores.  Signals like this are
also problematic for the \naive ~prediction strategy, since the
outliers have significant influence on the mean.  This compounds the
effect of the skew in the scaling factor and exacerbates the spread in
the \svdone MASE values.  For all of these reasons, we view \svdone
as an outlier and exclude it from the fit calculation of the dashed
line in Figure~\ref{fig:wpe_vs_mase_best}.

>>>>

In this section, we offer an empirical validation of the two
conjectures offered in Section \ref{sec:intro}, namely:

@@

\item That the weighted permutation entropy (WPE) of a noisy
  real-valued time series is correlated with prediction
  accuracy---i.e., that the predictable structure in a time-series
  data set can be quantified by its WPE.

@@

\item That the way information is generated and processed by a system
  is correlated with the effectiveness of a given predictor on
  time-series data from that system

@@

Figure~\ref{fig:wpe_vs_mase_best} shows the \emph{best} of these 45
predictions for each system: that is, the lowest error over all 15
trials and all three methods for each of the eight programs.  The WPE
is plotted against the corresponding MASE value here in order to bring
out the correlation between these two quantities.
\begin{figure}[htbp]
  \centering
  \includegraphics[width=\columnwidth]{figs/prediction_vs_entropy}
  \caption{Weighted permutation entropy vs. mean absolute scaled error
    (MASE) of the best prediction of each system.
% 
% (across all trials and all prediction methods) 
% 
The dashed line is a least-squares linear fit of all the points except
{\tt dgesdd$_1$}, which we have excluded for reasons explained in the
text.}
  \label{fig:wpe_vs_mase_best}
\end{figure}
As the plot shows, the best-case prediction error for each of these
eight systems is roughly proportional to the weighted permutation
entropy, which is consistent with our first conjecture.  The dashed
line in the figure, a linear least-squares fit to these best-case
errors\footnote{with the exception of {\tt dgesdd$_1$}, for reasons
  described below} captures this rough proportionality.  This is not a
formal result.  The three forecast methods used here were chosen to
span the space of standard prediction strategies, but they do not
cover that space exhaustively.  Our goal here is an empirical
assessment of predictability and complexity, not formal results about
a ``best'' predictor for a given time series.  There may be other
methods that produce lower MASE values than those in
Figure~\ref{fig:wpe_vs_mase_best}, but the sparseness of the points in
the upper-left and lower-right quadrants of this plot strongly
suggests that the underlying predictability of a time series is
inversely correlated with its WPE.  The rest of this section describes
our results in more detail---including the measures taken to assure
meaningful comparisons across methods, trials, and programs---and
elaborates on the meaning of the dashed line in the figure.

Figure~\ref{fig:wpe_vs_mase_all} shows WPE vs. MASE plots for the full
set of experiments.
\begin{figure}
  \centering
%    \includegraphics[width=\columnwidth]{figs/ARIMA_prediction_vs_entropy}
%    \includegraphics[width=\columnwidth]{figs/ARIMA_prediction_vs_entropy}
%    \includegraphics[width=\columnwidth]{figs/LMA_prediction_vs_entropy}
  \includegraphics[width=\columnwidth]{figs/predictions_vs_entropy3}
\caption{WPE vs. MASE for all trials, methods, and systems---with the
  exception of \svdone, \svdthree, and \svdfive are omitted from the
  top plot for scale reasons.
% 
% ; their MASE scores are $2.676 \pm 4.328$, $31.39 \pm 0.282$, and
%   $20.87 \pm 0.192$ respectively. 
% 
Numerical values, including means and standard deviations of the
errors, can be found in Table~\ref{tab:error}.  The dashed line is the
same as in the previous figure.
}
    \label{fig:wpe_vs_mase_all}
\end{figure} 
There are 15 points in each cluster, one for each trial.  (The points
in Figure~\ref{fig:wpe_vs_mase_best} are the leftmost of the points
for the corresponding system in any of the three plots in
Figure~\ref{fig:wpe_vs_mase_all}.)  The WPE values do not vary very
much across trials.  For most traces, the variance in MASE scores is
low as well, resulting in small, tight clusters.  In some
cases---ARIMA predictions of {\tt col\_major}, for instance---the MASE
variance is larger, which spreads out the clusters horizontally.  The
distribution of the cluster \emph{positions} is quite different in the
three plots.  The mean MASE scores of predictions generated with the
nonlinear LMA method are generally closer to the dashed line; the
ARIMA method clusters are more widely spread, the \naive ~clusters
even more so.  This geometric validation of the second conjecture in
this paper is discussed in more detail in the following paragraphs.

@@

The dashed line in Figures~\ref{fig:wpe_vs_mase_best}
and~\ref{fig:wpe_vs_mase_all} is a heuristic that captures the
argument at the end of the previous paragraph.  It can allow
practitioners to recognize when a prediction method is not well
matched to the task at hand: that is, when the time series has more
predictive structure than the method is able to use.  When the
predictions are well below that line (as in the top and middle images
in Figure~\ref{fig:wpe_vs_mase_all}), one should try a different
method.  Conversely, the position of the LMA cluster---well to the
left of the other two, near the dashed line---reflects this method's
ability to capture and exploit the structure that is present in this
signal.

@@

\svdone---the dark blue segment of
Figure~\ref{fig:svd-ts-colored}---behaves very differently than the
other seven systems in this study.  Though its weighted permutation
entropy is very high ($0.957 \pm 0.016$), two of the three prediction
methods do quite well on this signal, yielding MASE scores of $0.827
\pm 0.076$ (LMA) and $0.714 \pm 0.075$ (ARIMA).  This pushes the
corresponding clusters of points in the bottom two images in
Figure~\ref{fig:wpe_vs_mase_all} well above the trend followed by the
other seven signals.  The MASE scores of the predictions that were
produced by the \naive ~method for this system, however, are highly
inconsistent.  The majority of the blue diamond-shaped points on the
top plot in Figure~\ref{fig:wpe_vs_mase_all} are clustered near a MASE
score of 0.6, which is better than LMA and ARIMA.  In five of the 15
\svdone trials, however, there were step changes in the signal.  The
\naive ~method has a very difficult time with signals like this,
particularly if there are multiple step changes.  This raised the MASE
scores of these trials, pushing the corresponding points to the
right\footnote{This includes the cluster of three points near MASE
  $\approx 2.5$, as well as two points that are beyond the domain of
  the graph, at MASE 11.2 and 14.8.}, and in turn raising both the
mean and variance of this set of trials.

The effects described in the previous paragraph are exacerbated by the
way MASE is calculated.  Recall that MASE scores are scaled
\emph{relative to a random-walk forecast}, which uses the current
value as the prediction.  This strategy works very badly on signals
with frequent, large, rapid transitions.  Consider a signal that
oscillates from one end of its range to the other at every step.  A
signal like this will have a low WPE, much like {\tt col\_major}.
However, a random-walk forecast of this signal will be 180 degrees out
of phase with the true continuation.  Since random-walk error appears
in the denominator of the MASE score, this effect can shift points
leftwards on a WPE vs. MASE plot, and that is exactly why the
\svdone clusters for ARIMA and LMA are above the dashed line in
Figure~\ref{fig:wpe_vs_mase_all}.  This time series, which is shown in
closeup in Figure~\ref{fig:svdone-ts},
\begin{figure}[htbp]
  \centering
    \includegraphics[width=\columnwidth]{figs/svdonets2}
\caption{A small portion of the \svdone time series}\label{fig:svdone-ts}
\end{figure} 
is not quite to the level of the worst-case signal described above,
but it still poses a serious challenge to random-walk prediction.  It
is dominated by a noisy regime (between $\approx$1.86 and
$\approx$1.88 on the vertical scale in Figure~\ref{fig:svdone-ts}),
punctuated by short excursions above 1.9.  In the former regime, which
makes up more than 80\% of the signal, there are frequent dips to 1.82
and occasional larger dips below 1.8.  These single-point dips are the
bane of random-walk forecasting.  In this particular case, roughly
40\% of the forecasted points are off by the width of the associated
dip, which skews the associated MASE scores.  Signals like this are
also problematic for the \naive ~prediction strategy, since the
outliers have significant influence on the mean.  This compounds the
effect of the skew in the scaling factor and exacerbates the spread in
the \svdone MASE values.  For all of these reasons, we view \svdone as
an outlier and exclude it from the fit calculation of the dashed line
in Figure~\ref{fig:wpe_vs_mase_best}.

### 2014-04-22 wpe.tex
Liz is done rewriting

For the purposes of this paper, we view the Shannon entropy---in
particular its growth rate with respect to word length (the
\emph{Shannon entropy rate})---as a measure of complexity and
unpredictability in a time series.  Time-series consisting of i.i.d.
random variables, such as white noise, have maximal entropy rates, 
whereas highly structured time-series, for example periodic, have very
low or zero entropy rate. A time series with a high entropy rate is 
almost completely unpredictable; conversely, one with low entropy rate
is often quite predictable. This can be made more rigorous: Pesin's
relation \cite{pesin77} states that in chaotic dynamical systems, the
Kolmogorov-Sinai (KS) entropy is equal to the sum of the positive
Lyapunov exponents, $\lambda_i$.  The Lyapunov exponents directly
quantify the rate at which nearby states of the system diverge with
time: $\left| \Delta x(t) \right| \approx e^{\lambda t} \left| \Delta
x(0) \right|$.  The faster the divergence, the larger the entropy, and
the more difficult prediction becomes.  The KS entropy is defined as
the supremum of the Shannon entropy rates of all
partitions~\cite{petersen1989}. The partition that achieves this
supremum is the \emph{generating partition} discussed in
Section~\ref{sec:related}.

From a different point of view, we can consider the information (as
measured by the Shannon entropy) contained in a single observable, say
the \emph{present}, of the system. This information can be partitioned
into two components: the information shared with past
observations---e.g., the mutual information between the past and
present---and the information in the present that is not contained in
the past (aka ``the conditional entropy of the present given the
past'').  The first part is known as the \emph{redundancy}, as it is
information in the present that is also in the past.  The second part
is the aforementioned \emph{Shannon entropy rate}.  We conjecture that 
the more redundancy in a signal, the more predictable it is.  
Redundancy is critical to prediction, in general.  And the specific 
\emph{form} of the redundancy dictates whether a particular prediction 
method will work well or poorly on the corresponding signal.  A linear 
method cannot detect or make use of nonlinear redundancy, for 
instance.  This issue is central to the claims in this paper and the 
discussion in the following section.

Previous approaches to measuring temporal complexity via the Shannon
entropy rate \cite{Shannon1951, mantegna1994linguistic} required
categorical data: $x_i \in \mathcal{S}$ for some finite or countably
infinite \emph{alphabet} $\mathcal{S}$.  Data taken from real-world
systems is, however, effectively\footnote{Measurements from
  finite-precision sensors are discrete, but data from modern
  high-resolution sensors are, for the purposes of entropy
  calculations, effectively continuous.}  real-valued.  To analyze
real-valued data using a method that requires categorical values, one
must discretize the data---typically by binning.  Unfortunately, this
is rarely a good solution, as the binning of the values introduces
spurious dynamics~\cite{bollt2001}.  The field of symbolic dynamics
offers discretization methods that do not disturb the intrinsic
behavior.  These methods are, however, fragile in the face of noise;
worse yet, they require knowledge of the underlying system.  This is
not useful in our context, where we want to measure the entropy to
quantify predictability of an arbitrary time-series.

Bandt and Pompe introduced the \emph{permutation entropy} (PE) as a
``natural complexity measure for time series''
\cite{bandt2002per}. The permutation entropy employs a method of
discretizing real-valued time series that follows the intrinsic
behavior of the system under examination, and has many advantages 
including being robust to observational noise and its application does 
not require any knowledge of the underlying mechanisms.  Rather than 
looking at the statistics of sequences of values, as is done when 
computing the Shannon entropy, permutation entropy looks at the 
statistics of the \emph{orderings} of sequences of values using 
ordinal analysis. Ordinal analysis of a time series is the process of 
mapping successive time-ordered elements of a time series to their
value-ordered permutation of the same size.  By way of example, if
$(x_1, x_2, x_3) = (9, 1, 7)$ then its \emph{ordinal pattern},
$\phi(x_1, x_2, x_3)$, is $231$ since $x_2 \leq x_3 \leq x_1$.  The
ordinal pattern of the permutation $(x_1, x_2, x_3) = (9, 7, 1)$ is
$321$.

@@

Notice that $0\le H(\ell) \le \log_2(\ell!)$ \cite{bandt2002per}.
With this in mind, it is common in the literature to normalize
permutation entropy as follows: $\frac{H(\ell)}{\log_2(\ell!)}$.  With
this convention, ``low'' PE is close to 0 and ``high'' PE is close to
1. Finally, it should be noted that the permutation entropy has been
shown to be identical to the Kolmolgorov-Sinai entropy for many large
classes of systems \cite{amigo2012permutation}, so long as
observational noise is sufficiently small. As mentioned before, this
is equal to the Shannon entropy rate of a generating partition of the
system. This transitive chain of equalities, from permutation entropy
to Shannon entropy rate via the KS entropy, then allows us to
approximate the redundancy---being the dual of the Shannon entropy
rate---of a signal by $1 - \frac{H(\ell)}{\log_2(\ell!)}$.

@@

As a final note: there has been prior work under a similar title to
ours~\cite{haven2005}, but there are only superficial similarities
between the two research projects.  Haven {\sl et al.} utilize the
relative entropy to quantify the difference in predictability between 
two distributions: one using the known dynamical system from a small 
ensamble in the past, and the other the asymptotic distribution. Our 
work quantifies the predictability of a single observed time series 
using weighted permutation entropy and makes no assumptions about the
generating process.

>>>>

For the purposes of this paper, we view the Shannon entropy---in
particular its growth rate with respect to word length (the
\emph{Shannon entropy rate})---as a measure of complexity and
unpredictability in a time series.  Time-series consisting of i.i.d.
random variables, such as white noise, have maximal entropy rates,
whereas highly structured time-series, for example periodic, have very
low (or zero) entropy rates. A time series with a high entropy rate is
almost completely unpredictable; conversely, one with low entropy rate
is often quite predictable. This can be made more rigorous: Pesin's
relation \cite{pesin77} states that in chaotic dynamical systems, the
Kolmogorov-Sinai (KS) entropy is equal to the sum of the positive
Lyapunov exponents, $\lambda_i$.  The Lyapunov exponents directly
quantify the rate at which nearby states of the system diverge with
time: $\left| \Delta x(t) \right| \approx e^{\lambda t} \left| \Delta
x(0) \right|$.  The faster the divergence, the larger the entropy.
The KS entropy is defined as the supremum of the Shannon entropy rates
of all partitions~\cite{petersen1989}. The partition that achieves
this supremum is the \emph{generating partition} discussed in
Section~\ref{sec:related}.

From a different point of view, we can consider the information (as
measured by the Shannon entropy) contained in a single observable, say
the \emph{present}, of the system. This information can be partitioned
into two components: the information shared with past
observations---e.g., the mutual information between the past and
present---and the information in the present that is not contained in
the past (aka ``the conditional entropy of the present given the
past'').  The first part is known as the \emph{redundancy}, as it is
information in the present that is also in the past.  The second part
is the aforementioned \emph{Shannon entropy rate}.  We conjecture that 
the more redundancy in a signal, the more predictable it is.  
Redundancy is critical to prediction, in general.  And the specific 
\emph{form} of the redundancy dictates whether a particular prediction 
method will work well or poorly on the corresponding signal.  A linear 
method cannot capture or make use of nonlinear redundancy, for 
instance.  This issue is central to the claims in this paper and the 
discussion in the following section.

Previous approaches to measuring temporal complexity via the Shannon
entropy rate \cite{Shannon1951, mantegna1994linguistic} required
categorical data: $x_i \in \mathcal{S}$ for some finite or countably
infinite \emph{alphabet} $\mathcal{S}$.  Data taken from real-world
systems are, however, effectively\footnote{Measurements from
  finite-precision sensors are discrete, but data from modern
  high-resolution sensors are, for the purposes of entropy
  calculations, effectively continuous.}  real-valued.  To analyze
real-valued data using a method that requires categorical values, one
must discretize the data---typically by binning.  Unfortunately, this
is rarely a good solution, as the binning of the values introduces
spurious dynamics~\cite{bollt2001}.  The field of symbolic dynamics
offers discretization methods that do not disturb the intrinsic
behavior.  These methods are, however, fragile in the face of noise;
worse yet, they require knowledge of the underlying system.  This is
not useful in our context, where we want to measure the entropy to
quantify predictability of an arbitrary time-series.

Bandt and Pompe introduced the \emph{permutation entropy} (PE) as a
``natural complexity measure for time series''
\cite{bandt2002per}. The permutation entropy employs a method of
discretizing real-valued time series that follows the intrinsic
behavior of the system under examination.  It has many advantages,
including being robust to observational noise, and its application
does not require any knowledge of the underlying mechanisms.  Rather
than looking at the statistics of sequences of values, as is done when
computing the Shannon entropy, permutation entropy looks at the
statistics of the \emph{orderings} of sequences of values using
ordinal analysis. Ordinal analysis of a time series is the process of
mapping successive time-ordered elements of a time series to their
value-ordered permutation of the same size.  By way of example, if
$(x_1, x_2, x_3) = (9, 1, 7)$ then its \emph{ordinal pattern},
$\phi(x_1, x_2, x_3)$, is $231$ since $x_2 \leq x_3 \leq x_1$.  The
ordinal pattern of the permutation $(x_1, x_2, x_3) = (9, 7, 1)$ is
$321$.

@@

Notice that $0\le H(\ell) \le \log_2(\ell!)$ \cite{bandt2002per}.
With this in mind, it is common in the literature to normalize
permutation entropy as follows: $\frac{H(\ell)}{\log_2(\ell!)}$.  With
this convention, ``low'' PE is close to 0 and ``high'' PE is close to
1. Finally, it should be noted that the permutation entropy has been
shown to be identical to the Kolmolgorov-Sinai entropy for many large
classes of systems \cite{amigo2012permutation}, as long as
observational noise is sufficiently small. As mentioned before, this
is equal to the Shannon entropy rate of a generating partition of the
system. This transitive chain of equalities, from permutation entropy
to Shannon entropy rate via the KS entropy, allows us to approximate
the redundancy---being the dual of the Shannon entropy rate---of a
signal by $1 - \frac{H(\ell)}{\log_2(\ell!)}$.

@@

As a final note: there has been prior work under a similar title to
ours~\cite{haven2005}, but there are only superficial similarities
between the two research projects.  Haven \emph{et al.} utilize the
relative entropy to quantify the difference in predictability between
two distributions: one {\color{red}[[Ryan: a distribution ``using'' a
      dynamical system is confusing.  Do you mean ``created using''?
      Or ``measured from''?  And what do you mean by ``from an
      ensemble''?  ensemble of what?  used how?]]}  using the known
dynamical system from a small ensemble in the past, and the other the
asymptotic distribution. Our work quantifies the predictability of a
single observed time series using weighted permutation entropy and
makes no assumptions about the generating process.
